# Tutorial 4: Inference and tests against zero {#b-ep04}

```{r bep04-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
pitch_sample <- sharks[sample(seq_len(nrow(sharks)), 100), ]
pitch_sample$equity_gap_pp <-
  pitch_sample$equity_offered_pct - 15
```

<div class="outcomes">
<span class="label">By the end of Tutorial 4, you can</span>
<ul>
<li>construct and interpret intervals for a mean and proportion;</li>
<li>complete two-sided tests of zero-valued contrasts;</li>
<li>connect a test decision with its confidence interval;</li>
<li>interpret degrees of freedom, errors, and power;</li>
<li>read a regression coefficient test without overclaiming.</li>
</ul>
</div>

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 9 and 10. Review
[Mathematics Refresher: intervals, degrees of freedom, and
standardisation](#math-refresher) if needed.

Use the cumulative [formula sheet through Lecture 10](#formula-l10).

Write from memory:

1. the t confidence-interval form;
2. the t statistic for testing a mean contrast against zero;
3. the meaning of \(df=n-1\);
4. the definition of a two-sided p-value;
5. the 95%-interval/5%-test decision link.

<details>
<summary>Check your preparation</summary>

1. \(\widehat\delta\pm t_{0.975,n-1}
   \operatorname{SE}(\widehat\delta)\).
2. \(T=\widehat\delta/\operatorname{SE}(\widehat\delta)\).
3. Estimating the sample mean leaves \(n-1\) freely varying deviations for
   estimating spread.
4. Under \(H_0\), the probability of a statistic at least as far from zero as
   the observed statistic in either direction.
5. For matched two-sided t procedures, reject zero exactly when the 95%
   interval excludes zero.

</details>

:::

::: {.course-phase .phase-in-class}

## In class

Attempt every task before opening its answer. Every conclusion must name the
parameter, units, population, direction, and evidence limit.

### Task 1 — Mean contrast

For the reproducible sample, define

\[
D=\text{initial equity offered}-15.
\]

1. Calculate \(\bar d\), \(s_d\), the SE, and degrees of freedom.
2. Construct a 95% interval for \(\delta=E[D]\).
3. Test \(H_0:\delta=0\) against \(H_1:\delta\ne0\).
4. Explain why the interval and test agree.

<details>
<summary>Check your answer</summary>

```{r bep04-mean-gap}
d <- pitch_sample$equity_gap_pp
n <- length(d)
d_bar <- mean(d)
s_d <- sd(d)
se_d <- s_d / sqrt(n)
critical <- qt(0.975, df = n - 1)
ci_d <- d_bar + c(-1, 1) * critical * se_d
t_value <- d_bar / se_d
p_value <- 2 * pt(-abs(t_value), df = n - 1)

c(estimate = d_bar, sd = s_d, se = se_d, df = n - 1,
  lower = ci_d[1], upper = ci_d[2],
  t = t_value, p_value = p_value)
```

\(\bar d=-2.24\), \(s_d=7.7079\), \(SE=0.7708\), and \(df=99\).
The 95% interval is approximately \((-3.77,-0.71)\). The statistic is
\(t=-2.906\) and the two-sided p-value about 0.0046, so reject the zero-gap
null. The interval excludes zero because it uses the same t model and SE as
the test.

</details>

### Task 2 — Proportion contrast

Define \(\delta_p=p-0.50\), where \(p\) is the finite-frame deal proportion.

1. Estimate \(\delta_p\).
2. Perform a two-sided large-sample z-test against zero.
3. Construct the course's large-sample 95% interval for \(p\) and translate it
   into an interval for \(\delta_p\).
4. State why test and interval SEs differ.

<details>
<summary>Check your answer</summary>

```{r bep04-proportion-gap}
n <- nrow(pitch_sample)
phat <- mean(pitch_sample$deal_on_show)
p0 <- 0.50
delta_hat <- phat - p0

null_se <- sqrt(p0 * (1 - p0) / n)
z_value <- delta_hat / null_se
p_value <- 2 * pnorm(-abs(z_value))

estimated_se <- sqrt(phat * (1 - phat) / n)
ci_p <- phat + c(-1, 1) * 1.96 * estimated_se
ci_delta <- ci_p - p0

c(phat = phat, estimated_gap = delta_hat,
  z = z_value, p_value = p_value)
c(proportion_lower = ci_p[1], proportion_upper = ci_p[2],
  gap_lower = ci_delta[1], gap_upper = ci_delta[2])
```

\(\hat p=0.64\), so \(\widehat\delta_p=0.14\). Under the null,
\(SE_0=0.05\), \(z=2.80\), and the two-sided p-value is about 0.0051.
Reject the zero-gap null. The proportion interval is approximately
\((0.546,0.734)\), corresponding to a gap interval \((0.046,0.234)\).
The test calculates variation under \(p_0\); the manual interval estimates it
with \(\hat p\).

</details>

### Task 3 — Read a regression coefficient row

Suppose a model of initial equity offered reports:

| Predictor | Estimate | SE | t | p-value | 95% interval |
|---|---:|---:|---:|---:|---:|
| Amount requested, per $100,000 | -0.290 | 0.058 | -5.00 | <0.001 | (-0.403, -0.177) |

1. State the tested hypotheses.
2. Explain every number in units.
3. Give the statistical conclusion at 5%.
4. State two claims the row cannot justify.

<details>
<summary>Check your answer</summary>

The test is \(H_0:\beta_{ask}=0\) against \(H_1:\beta_{ask}\ne0\).
The estimate says that another $100,000 requested is associated with 0.290
percentage points less fitted initial equity, holding the model's other
predictors fixed. The SE measures sampling uncertainty in the same coefficient
units. The statistic is \(-0.290/0.058=-5.00\). The p-value is below 0.001,
and the 95% interval excludes zero, so reject the zero-coefficient null. The
row alone does not establish causality or show that model assumptions hold.

</details>

### Task 4 — Errors and power

For Task 3:

1. describe a Type I error;
2. describe a Type II error;
3. state how a larger sample affects SE and power;
4. explain why raising \(\alpha\) is a trade-off rather than a free gain.

<details>
<summary>Check your answer</summary>

A Type I error is concluding that the population-model ask coefficient differs
from zero when it equals zero. A Type II error is failing to reject zero when a
specified non-zero coefficient is true. A larger independent sample usually
reduces SE and raises power. Raising \(\alpha\) also raises power, but it
increases the Type I error probability.

</details>

### Task 5 — Repair the conclusion

Critique:

> The coefficient is significant, so asking for more money causes founders to
> offer less equity, and there is less than a 0.1% chance that the null is true.

<details>
<summary>Check your answer</summary>

The p-value is not the probability that the null is true, and significance
does not establish causality. A defensible conclusion reports a negative
adjusted association, its magnitude and interval, the evidence against a zero
coefficient under the model, and the observational design and assumption
limits.

</details>

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice — 25 minutes

A random sample of 81 customers gives mean spending 42 euros, sample SD 9
euros, and 54 renewals.

1. Define \(\delta_\mu=\mu-40\), construct its 95% interval, and test a
   two-sided zero null.
2. Define \(\delta_p=p-0.60\), construct the course's 95% interval, and test a
   two-sided zero null.
3. State the Type I error for each test.
4. Explain why the mean result is only just significant while the estimated
   mean difference is still 2 euros.
5. Write a six-sentence report separating estimate, uncertainty, decision,
   practical importance, population, and causality.

<details>
<summary>Check a model answer</summary>

For the mean, \(\widehat\delta_\mu=2\), \(SE=9/\sqrt{81}=1\), and
\(df=80\). With \(t_{0.975,80}\approx1.990\), the gap interval is
approximately \((0.01,3.99)\), equivalent to a mean interval
\((40.01,43.99)\). The statistic is \(t=2.00\) and the two-sided p-value is
about 0.049, so reject the zero mean-gap null just narrowly.

For renewal, \(\hat p=54/81=0.667\). The estimated gap is 0.067. The
large-sample proportion interval is approximately \((0.564,0.769)\), so the
gap interval is \((-0.036,0.169)\). Under \(H_0:p=0.60\),
\(SE_0\approx0.0544\), \(z\approx1.225\), and the two-sided p-value is about
0.221; do not reject.

A Type I error would be claiming a non-zero mean spending gap when
\(\mu=40\), or a non-zero renewal gap when \(p=0.60\). The observed estimate
does not change when \(\alpha\) changes; only the decision threshold does. A
complete report must still evaluate whether a 2-euro difference matters and
whether the sampling design supports the intended population claim. Neither
test identifies a causal explanation.

</details>

QM Basic is complete. Next, use the [QM Basic Formula Sheet](#formula-sheet)
and attempt the [QM Basic Mock Exam](#basic-mock-exam). Continue to
[QM Intermediate](#intermediate-course) when that course begins.

:::
