# Tutorial 2: Conditioning and relationships {#b-ep02}

```{r bep02-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
L <- sharks$season >= 5
```

<div class="outcomes">
<span class="label">By the end of Tutorial 2, you can</span>
<ul>
<li>calculate variance as an expected squared deviation;</li>
<li>calculate and interpret conditional and joint probabilities;</li>
<li>apply independence, total probability, and Bayes' rule;</li>
<li>recover marginal distributions from a joint PMF;</li>
<li>interpret covariance and correlation without making causal claims.</li>
</ul>
</div>

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 4 and 5. Complete these statements
without reopening the chapters:

1. In \(P(A\mid B)\), the reference class is ...
2. Variance averages ...
3. Events \(A\) and \(B\) are independent when ...
4. A joint PMF records ...
5. Covariance averages ...

<details>
<summary>Check your preparation</summary>

1. The reference class is \(B\), the event after the vertical bar.
2. Variance averages squared deviations from a variable's mean.
3. \(P(A\cap B)=P(A)P(B)\).
4. A joint PMF records the probability of every possible value pair.
5. Covariance averages products of deviations from the two means.

</details>

:::

::: {.course-phase .phase-in-class}

## In class

Attempt each task before opening its answer. Use the formula and reference
class, verify the value, interpret it in context, and qualify the claim.

### Task 1 — Conditional probability and Bayes

```{r bep02-table, echo=FALSE}
tab <- table(
  Period = factor(L, levels = c(FALSE, TRUE),
                  labels = c("Seasons 1-4", "Seasons 5-8")),
  Deal = factor(D, levels = c(FALSE, TRUE),
                labels = c("No on-air deal", "On-air deal"))
)
tab
```

1. Calculate and interpret \(P(D\mid L)\).
2. Calculate and interpret \(P(L\mid D)\).
3. Verify \(P(D\cap L)=P(D\mid L)P(L)\).
4. Use total probability across \(\{L^c,L\}\) to recover \(P(D)\).
5. Show how Bayes' rule recovers \(P(L\mid D)\).

<details>
<summary>Check your answer</summary>

\(P(D\mid L)=256/444=0.577\): among late-season records, 57.7% reached an
on-air deal. \(P(L\mid D)=256/383=0.668\): among deal records, 66.8% came from
late seasons. The joint probability is
\((256/444)(444/706)=256/706\).

\[
P(D)=\frac{127}{262}\frac{262}{706}
     +\frac{256}{444}\frac{444}{706}
     =\frac{383}{706}.
\]

Bayes gives
\[
P(L\mid D)=\frac{P(D\mid L)P(L)}{P(D)}
=\frac{(256/444)(444/706)}{383/706}
=\frac{256}{383}.
\]

</details>

### Task 2 — Independence

R reports:

```{r bep02-independence, echo=FALSE}
c(
  P_D = mean(D),
  P_D_given_L = mean(D[L]),
  P_D_and_L = mean(D & L),
  P_D_times_P_L = mean(D) * mean(L)
)
```

1. Use the conditional definition to assess empirical independence.
2. Use the product definition to reach the same conclusion.
3. Explain why “not independent” does not mean “late season causes a deal.”
4. Explain why the early and late events are disjoint but not independent.

<details>
<summary>Check your answer</summary>

The deal probability is about 0.542 overall and 0.577 within late seasons, so
conditioning changes it. Likewise, the observed joint probability is about
0.363 rather than the independent product of about 0.341. Thus the empirical
events are not independent. This is association in selected observational
records; a time-linked cause is not isolated. Early and late are disjoint
because they cannot occur together. Observing one makes the other impossible,
so positive-probability disjoint events are dependent.

</details>

### Task 3 — Joint PMF, covariance, and correlation

Let \(X\) be the 0/1 deal indicator and \(Y\) the 0/1 late-period indicator.

```{r bep02-joint, echo=FALSE}
joint <- prop.table(table(X = as.integer(D), Y = as.integer(L)))
joint

x <- as.integer(D)
y <- as.integer(L)
c(
  E_X = mean(x),
  E_Y = mean(y),
  Var_X = mean((x - mean(x))^2),
  Var_Y = mean((y - mean(y))^2),
  E_XY = mean(x * y),
  covariance = mean(x * y) - mean(x) * mean(y),
  correlation = cor(x, y)
)
```

1. Verify that the joint PMF sums to 1.
2. Recover the marginal PMF of \(X\) by summing rows.
3. Verify \(\operatorname{Var}(X)=E[X^2]-E[X]^2\).
4. Verify
   \(\operatorname{Cov}(X,Y)=E[XY]-E[X]E[Y]\).
5. Interpret the sign of covariance.
6. Explain what correlation changes and what it does not prove.

<details>
<summary>Check your answer</summary>

The four joint masses sum to 1. Summing across the two \(Y\) columns gives
\(P(X=0)=323/706\) and \(P(X=1)=383/706\). Because \(X\) is binary,
\(E[X^2]=E[X]\), so \(\operatorname{Var}(X)=E[X]-E[X]^2\). Here
\(E[XY]=P(X=1,Y=1)=256/706\); subtracting
\((383/706)(444/706)\) gives a positive covariance. Positive covariance means
the two 1-coded values co-occur slightly more than under the product of their
marginal means. Correlation divides covariance by both SDs, making the result
unitless. A zero correlation would not establish independence; correlation
does not establish causality or practical importance.

</details>

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice — 10 minutes

```{r bep02-timed-output, echo=FALSE}
c(
  early_deal_rate = mean(D[!L]),
  late_deal_rate = mean(D[L]),
  cor_season_words = cor(sharks$season, sharks$description_words),
  cor_words_deal = cor(sharks$description_words, sharks$deal_on_show)
)
```

Write no more than 120 words interpreting all four values. Include denominators,
direction and rough strength, measurement, and the causal boundary.

<details>
<summary>Check a model answer</summary>

Among Seasons 1--4, 48.5% of records reached an on-air deal; among Seasons
5--8, 57.7% did so. The 9.2-percentage-point difference is descriptive and does
not identify a causal effect of season. Season and written-description length
have a moderate positive linear association (\(r\approx0.43\)); description
length and the deal indicator have a weak positive linear association
(\(r\approx0.14\)). Word count measures dataset text, and the outcome is an
on-air agreement rather than completed investment.

</details>

### Check your understanding

Complete from memory:

1. The conditioning event supplies ...
2. Independence means ...
3. Covariance averages ...
4. Correlation standardises ...

<details>
<summary>Check your answers</summary>

1. The conditioning event supplies the denominator or reference class.
2. Events \(A\) and \(B\) are independent when
   \(P(A\cap B)=P(A)P(B)\).
3. Covariance averages products of deviations from the two means.
4. Correlation standardises covariance by dividing by both standard
   deviations.

</details>

The next theory chapter, Lecture 6, extends the CDF framework to continuous random
variables and introduces PDFs, uniform distributions, and normal distributions.

:::
