# Lecture 2: Probability foundations and random variables {#b02}

```{r b02-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
```

<div class="outcomes">
<span class="label">By the end of Lecture 2, you can</span>
<ul>
<li>read frequency and relative-frequency distributions;</li>
<li>choose a bar chart or histogram and describe centre, spread, and shape;</li>
<li>define a random experiment, outcome, sample space, and event;</li>
<li>use complements, intersections, unions, and disjointness;</li>
<li>state the probability axioms and distinguish empirical proportions from model probabilities;</li>
<li>distinguish an outcome, a random variable, and a realised value.</li>
</ul>
</div>

::: {.course-phase .phase-before}

## Before class

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch <a href="https://www.youtube.com/watch?v=iQ2edOqEQAs">Sample Space</a>
and <a href="https://www.youtube.com/watch?v=pA83XtLeVig">Probability
Axioms</a>, then watch
<a href="https://www.youtube.com/watch?v=vfqPpai_9jI">Definition of Random
Variables</a>. Be ready to define an outcome, sample space, event, and
random variable and to state the three probability axioms. All links are also
listed in the
<a href="videos.html#videos">External Video Guide</a>.
</div>

If set notation, complements, unions, or intersections are unfamiliar, review
[Mathematics Refresher: sets and events](#math-refresher).

An **empirical distribution** records how often values occur in a dataset.
For a categorical variable, a frequency table gives counts and a
relative-frequency table gives proportions.

```{r b02-season-table}
season_counts <- table(sharks$season)
season_counts
round(prop.table(season_counts), 3)
```

```{r b02-season-plot, echo=FALSE}
barplot(
  season_counts,
  col = "#0072B2", border = NA,
  bty = "n", las = 1,
  ylim = c(0, 1.08 * max(season_counts)),
  xlab = "Season", ylab = "Number of recorded pitches"
)
```

The horizontal axis lists the discrete season categories; each bar's height on
the vertical count scale is the number of recorded pitches in that season.
The later seasons contain more records, so an overall percentage gives those
seasons more weight.

For a quantitative variable, a histogram groups nearby values into intervals.

```{r b02-word-histogram, echo=FALSE}
hist(
  sharks$description_words,
  breaks = seq(0.5, 13.5, by = 1),
  col = "#E6F4F3", border = "white",
  main = "",
  xlab = "Words in the written dataset description",
  ylab = "Number of recorded pitches",
  xaxt = "n"
)
axis(1, at = 1:13)
abline(v = median(sharks$description_words), col = "#007C83", lwd = 3)
```

The horizontal axis measures description length in words and the vertical axis
counts pitches in each one-word interval. The teal vertical line marks the
median. The distribution is mildly right-skewed: most descriptions contain
3--7 words, with a small number extending farther to the right. The median is
5 words.

If needed, review [summary statistics](https://walshc.github.io/ebi-prog/dataframes-summary-statistics.html)
and [plots](https://walshc.github.io/ebi-prog/intro-to-plotting.html) from
Programming for E&BI.

:::

::: {.course-phase .phase-in-class}

## In class

### A probability model begins with possible outcomes

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/sample-space/">01.2 Sample Space</a>.</p>

Probability starts by specifying exactly what may happen. A probability model
needs a process, a list of complete possible results, and events that describe
which results matter for the question.

A **random experiment** is a process whose outcome is not known in advance.
The word *random* does not mean "unstructured." It means the model does not
specify in advance which possible outcome will occur.

- An **outcome**, written \(\omega\), is one complete possible result.
- The **sample space**, written \(\Omega\), is the set of all possible outcomes.
- An **event** is a specified subset of \(\Omega\) to which the model assigns a
  probability. It occurs when the realised outcome belongs to that subset.

The distinction is easiest to read from the whole-to-part structure:

| Object | Question it answers | Shark Tank example |
|---|---|---|
| experiment | What is done? | Select one of the 1,441 records uniformly |
| outcome \(\omega\) | Which complete result occurred? | One complete selected pitch record |
| sample space \(\Omega\) | Which complete results were possible? | The set of all 1,441 records |
| event \(A\) | Which outcomes count as “yes” for this question? | The subset of records with an on-air deal |

For a single roll of a fair six-sided die,

\[
\Omega=\{1,2,3,4,5,6\}.
\]

The event \(E=\{2,4,6\}\) means "an even result." The outcome \(\omega=4\)
belongs to \(E\), so \(E\) occurs when 4 is rolled. The number 4 is an outcome;
the singleton set \(\{4\}\) is an event.

#### An empirical selection experiment

Define the experiment as **select one of the 1,441 recorded pitches uniformly
at random**. The sample space is the set of 1,441 pitch records. Define:

- \(D\): the selected pitch reached an on-air deal;
- \(L\): the selected pitch appeared in Seasons 9--16.

Making the selection uniform means every record has probability \(1/1441\).
Consequently, event probabilities equal relative frequencies in this fixed
dataset. This does not imply the same probabilities for future entrepreneurs.

#### Event language is set language

| Notation | Name | Meaning |
|---|---|---|
| \(A^c\) | complement | \(A\) does not occur |
| \(A\cap B\) | intersection | both \(A\) and \(B\) occur |
| \(A\cup B\) | union | \(A\), \(B\), or both occur |
| \(A\subseteq B\) | subset | every outcome in \(A\) is also in \(B\) |
| \(A\cap B=\varnothing\) | disjoint | \(A\) and \(B\) cannot occur together |

```{r b02-event-diagrams, echo=FALSE, fig.width=7, fig.height=2.5}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3), mar = c(0.5, 0.5, 2.1, 0.5), xaxs = "i", yaxs = "i")

draw_frame <- function(title) {
  plot(NA, xlim = c(0, 10), ylim = c(0, 7), axes = FALSE,
       xlab = "", ylab = "", main = title, asp = 1)
  rect(0.3, 0.3, 9.7, 6.7, border = "#17324D", lwd = 2)
  text(9.1, 6.1, expression(Omega), col = "#17324D")
}

draw_frame("Complement: Aᶜ")
rect(0.3, 0.3, 9.7, 6.7, col = "#DCEAF3", border = NA)
rect(2.2, 1.5, 7.8, 5.5, col = "white", border = "#0072B2", lwd = 2)
rect(0.3, 0.3, 9.7, 6.7, border = "#17324D", lwd = 2)
text(5, 3.5, "A", col = "#0072B2")

draw_frame("Intersection: A ∩ B")
rect(1.4, 1.4, 6.2, 5.6, border = "#0072B2", lwd = 2)
rect(3.8, 1.4, 8.6, 5.6, border = "#D55E00", lwd = 2)
rect(3.8, 1.4, 6.2, 5.6, col = "#009E73", border = NA)
text(2.4, 3.5, "A", col = "#0072B2")
text(7.6, 3.5, "B", col = "#D55E00")

draw_frame(expression("Union: " * A %union% B))
rect(1.4, 1.4, 6.2, 5.6,
     col = grDevices::adjustcolor("#0072B2", alpha.f = 0.45),
     border = "#0072B2", lwd = 2)
rect(3.8, 1.4, 8.6, 5.6,
     col = grDevices::adjustcolor("#E69F00", alpha.f = 0.45),
     border = "#D55E00", lwd = 2)
text(2.4, 3.5, "A", col = "#0072B2")
text(7.6, 3.5, "B", col = "#D55E00")
par(old_par)
```

In each diagram, the outer rectangle is the sample space \(\Omega\), and the
coloured area is the event named above the panel. These are set diagrams: the
areas show which outcomes belong to an event, not the numerical size of its
probability.

Disjoint does **not** mean independent. If two disjoint events both have
positive probability, observing one makes the other impossible. Independence,
introduced in Lecture 5, means that observing one event does not change the
probability of the other.

### Probability laws and their axioms

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-axioms/">01.4 Probability Axioms</a>.</p>

A **probability law** assigns a number \(P(A)\) to each event \(A\). It must
satisfy:

1. **Non-negativity:** \(P(A)\ge 0\) for every event \(A\).
2. **Normalisation:** \(P(\Omega)=1\).
3. **Countable additivity:** for pairwise disjoint events
   \(A_1,A_2,\ldots\),

\[
P\left(\bigcup_{i=1}^{\infty} A_i\right)
=\sum_{i=1}^{\infty} P(A_i).
\]

The first two axioms set the probability scale: probability cannot be negative,
and something in the sample space must occur. The third says that mutually
exclusive routes can be added because no outcome is counted twice.

For a finite sample space with equally likely outcomes,

\[
P(A)=\frac{|A|}{|\Omega|}.
\]

That ratio is a special case, not a universal definition. It applies because
equal likelihood has been built into the model.

#### Rules derived from the axioms

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/simple-properties-of-probabilities/">01.5 Simple Properties of Probabilities</a>.</p>

Because \(A\) and \(A^c\) are disjoint and together form \(\Omega\),

\[
P(A^c)=1-P(A).
\]

When \(A\) and \(B\) may overlap,

\[
P(A\cup B)=P(A)+P(B)-P(A\cap B).
\]

The intersection is subtracted once because it was counted in both marginal
probabilities.

A **partition** divides \(\Omega\) into events that do not overlap and together
include every possible outcome. Each outcome belongs to exactly one part. For
example, Seasons 1--8 and Seasons 9--16 partition the recorded pitches into an
early-period event \(L^c\) and a late-period event \(L\).

```{r b02-partition-diagram, echo=FALSE, fig.width=6.4, fig.height=2.4}
old_par <- par(no.readonly = TRUE)
par(mar = c(0.3, 0.3, 0.3, 0.3))
plot(NA, xlim = c(0, 10), ylim = c(0, 5), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")
rect(0.4, 0.4, 9.6, 4.6, border = "#17324D", lwd = 2)
rect(0.4, 0.4, 4.9, 4.6, col = "#DCEAF3", border = NA)
rect(4.9, 0.4, 9.6, 4.6, col = "#FDE5C1", border = NA)
segments(4.9, 0.4, 4.9, 4.6, col = "#17324D", lwd = 2)
rect(0.4, 0.4, 9.6, 4.6, border = "#17324D", lwd = 2)
text(2.65, 2.5, expression(L^c), col = "#0072B2", cex = 1.25)
text(7.25, 2.5, "L", col = "#D55E00", cex = 1.25)
text(9.15, 4.15, expression(Omega), col = "#17324D")
par(old_par)
```

The two coloured regions do not overlap, and together they fill the whole
rectangle. Lecture 4 uses this idea to combine conditional probabilities from
separate groups.

```{r b02-probability-rules}
c(
  P_D = mean(D),
  P_L = mean(L),
  P_D_and_L = mean(D & L),
  P_not_D = mean(!D),
  P_D_or_L_direct = mean(D | L),
  P_D_or_L_rule = mean(D) + mean(L) - mean(D & L)
)
```

For example, \(P(D)=882/1441\approx0.612\) in the uniform record-selection
experiment.

For example, “no on-air deal” is \(D^c\), “late-season and deal” is
\(L\cap D\), and “late-season, deal, or both” is \(L\cup D\).

### Empirical proportion or future probability?

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/interpretations-uses-of-probabilities/">01.10 Interpretations and Uses of Probabilities</a>.</p>

The symbol \(P(D)\) is meaningful only after the probability model is defined.
For uniform selection from the file, it is exactly 882/1441. Calling it the
probability that a future televised pitch reaches an on-air agreement adds
assumptions:

- the future case belongs to a defined target population;
- selection into the show is comparable;
- the process is stable over time;
- "deal" retains the same definition.

Calling it the probability of completed investment would additionally change
the outcome being predicted; completed investment is not recorded here.

### From complete outcomes to numerical measurements

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/definition-of-random-variables/">05.2 Definition of Random Variables</a>.</p>

An outcome is a complete result, but an analysis usually focuses on one
numerical feature of that result. A **random variable** is the rule that extracts
that number. Formally,

\[
X:\Omega\rightarrow\mathbb{R}.
\]

Read the notation from left to right: \(X\) takes an outcome from the sample
space \(\Omega\) and returns a real number.

| Symbol | Meaning |
|---|---|
| \(\omega\) | one complete outcome, such as one selected pitch record |
| \(X\) | the fixed measurement rule |
| \(X(\omega)\) | the number assigned by \(X\) to outcome \(\omega\) |
| \(x\) | a possible or realised numerical value of \(X\) |

For example, define \(X(\omega)=1\) when pitch \(\omega\) reached an on-air
deal and \(X(\omega)=0\) otherwise. We could also define \(S(\omega)\) as its
season and \(W(\omega)\) as its written-description word count. The same
outcome therefore produces several measurements.

The rule \(X\) is defined before an outcome is selected. After selection,
\(x=X(\omega)\) is its **realised value**. The variable is called random
because its value is unknown before the outcome is observed, not because its
definition changes.

A random variable is **discrete** when its possible values form a finite or
countably infinite set. Its **support** is the set of values with positive
probability. The support of the deal indicator is \(\{0,1\}\), while the
support of season is \(\{1,\ldots,8\}\). Lecture 3 assigns probabilities to
these possible values.

:::

::: {.course-phase .phase-after}

## After class

### Retrieval

1. What is the difference between an outcome and an event?
2. What is the difference between a random variable and a realised value?
3. State the three probability axioms.
4. Why is the equally-likely ratio a special case?
5. What extra assumptions turn an empirical proportion into a forecast?

<details>
<summary>Check your answers</summary>

1. An outcome is one possible complete result; an event is a specified set of
   outcomes.
2. A random variable is a fixed function from outcomes to numbers; a realised
   value is the number produced for the observed outcome.
3. Probabilities are non-negative, \(P(\Omega)=1\), and probabilities add over
   any countable collection of pairwise disjoint events.
4. The ratio \(|A|/|\Omega|\) requires a finite sample space whose outcomes are
   equally likely.
5. A forecast requires a defined target population and assumptions about
   selection, process stability, and a stable outcome definition.

</details>

### Practice

In the uniform selection experiment, use
\(P(D)=882/1441\), \(P(L)=734/1441\), and
\(P(D\cap L)=505/1441\).

1. Calculate and interpret \(P(D^c)\).
2. Calculate and interpret \(P(D\cup L)\).
3. Are \(D\) and \(D^c\) disjoint? Explain.
4. Define a random variable \(S\) that records the season of the selected
   pitch. State its possible values.
5. Define a deal indicator \(X\). Explain how the event \(D\) and the random
   variable \(X\) are related but not the same object.

<details>
<summary>Check your answers</summary>

1. \(P(D^c)=1-882/1441=559/1441\approx0.388\): 38.8% of the records did not
   reach an on-air agreement.
2. \(P(D\cup L)=(882+734-505)/1441=1111/1441\approx0.771\): 77.1% of the
   records are from Seasons 9--16, reached an on-air deal, or satisfy both
   conditions.
3. They are disjoint because a record cannot simultaneously have and not have
   an on-air deal.
4. Define \(S(\omega)\) as the season number of pitch record \(\omega\).
   Its possible values are \(\{1,2,\ldots,16\}\).
5. Define \(X(\omega)=1\) when \(\omega\in D\) and \(X(\omega)=0\) otherwise.
   \(D\) is a set of outcomes; \(X\) is a function that assigns a number to
   every outcome. They are related by \(D=\{\omega:X(\omega)=1\}\).

</details>

### Worked exam interpretation

**Question.** R reports `mean(sharks$deal_on_show) = 0.612`. Is this an
empirical proportion or a universal probability?

<details>
<summary>Check a model answer</summary>

It is first an empirical proportion: 882 of the 1,441 recorded
pitches reached an on-air agreement. It is also the event probability in a
model that selects one of those 1,441 records uniformly. Applying it to a future
entrepreneur requires additional population, selection, stability, and outcome
assumptions.

</details>

Continue to [Lecture 3: Discrete distributions and expectation](#b03).

Formula sheet: [Lecture 2 formulas](#formula-l2).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 2--4 for empirical distributions and Chapters 6--8 for probability
models and random variables.

:::
