# Lecture 4: Variance, conditioning, and belief updating {#b04}

```{r b04-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
L <- sharks$season >= 5
p_hat <- mean(sharks$deal_on_show)
season_pmf <- prop.table(table(sharks$season))
s <- as.numeric(names(season_pmf))
p_s <- as.numeric(season_pmf)
```

<div class="outcomes">
<span class="label">By the end of Lecture 4, you can</span>
<ul>
<li>calculate and interpret variance and standard deviation;</li>
<li>distinguish distribution variance from the usual sample variance;</li>
<li>calculate conditional probabilities and identify their reference classes;</li>
<li>use total probability and Bayes' rule.</li>
</ul>
</div>

::: {.course-phase .phase-before}

## Before class

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch <a href="https://www.youtube.com/watch?v=ZWo1XgAQE5k">Variance</a>,
<a href="https://www.youtube.com/watch?v=MPRKc4UPoJk">Conditional
Probabilities</a>, and <a href="https://www.youtube.com/watch?v=kz2tvO_ZAKI">Bayes'
Rule</a>. Be ready to explain what variance averages, identify the reference
class in a conditional probability, and explain why reversing a condition
changes the denominator. All links are also listed in the
<a href="videos.html#videos">External Video Guide</a>.
</div>

Lecture 3 introduced expectation. For a random variable with mean
\(\mu=E[X]\), variance is the expected squared distance from that mean:

\[
\operatorname{Var}(X)=E[(X-\mu)^2].
\]

Before calculating it, explain why deviations are squared rather than simply
averaged.

<details>
<summary>Check your preparation</summary>

The probability-weighted deviations from the mean sum to zero. Squaring makes
both positive and negative deviations contribute to spread.

</details>

Compare:

- “Among pitches in Seasons 5-8, what proportion reached an on-air deal?”
- “Among pitches that reached an on-air deal, what proportion came from
  Seasons 5-8?”

Let \(D\) be the deal event and \(L\) the late-season event. The questions are
\(P(D\mid L)\) and \(P(L\mid D)\). The event after the vertical bar defines
the restricted reference class and therefore the denominator.

```{r b04-table, echo=FALSE}
period_deal <- table(
  Period = factor(L, levels = c(FALSE, TRUE),
                  labels = c("Seasons 1-4", "Seasons 5-8")),
  Deal = factor(D, levels = c(FALSE, TRUE),
                labels = c("No on-air deal", "On-air deal"))
)
period_deal
```

Before calculating, point to the numerator and denominator for:

1. \(P(D)\);
2. \(P(L)\);
3. \(P(D\cap L)\);
4. \(P(D\mid L)\);
5. \(P(L\mid D)\).

<details>
<summary>Check your denominators</summary>

1. \(P(D)\): 383 deal records out of 706 records.
2. \(P(L)\): 444 late-season records out of 706 records.
3. \(P(D\cap L)\): 256 late-season deal records out of 706 records.
4. \(P(D\mid L)\): 256 late-season deal records out of 444 late-season records.
5. \(P(L\mid D)\): 256 late-season deal records out of 383 deal records.

</details>

:::

::: {.course-phase .phase-in-class}

## In class

### Variance and standard deviation

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/variance/">06.2 Variance</a>.</p>

Expectation locates the centre of a distribution, but it does not describe how
tightly values cluster around that centre. Two distributions can have the same
expectation and very different spreads.

To measure spread, start with the deviation \(X-\mu\), where
\(\mu=E[X]\). Positive and negative deviations would cancel if they were
averaged directly, so variance squares each deviation before taking its
probability-weighted average. When \(E[X^2]<\infty\),

\[
\operatorname{Var}(X)
=E[(X-\mu)^2]
=\sum_x (x-\mu)^2p_X(x).
\]

Read the calculation as four steps:

1. locate the centre, \(\mu=E[X]\);
2. measure each possible deviation, \(x-\mu\);
3. square the deviation, \((x-\mu)^2\);
4. weight it by \(p_X(x)\) and add across the support.

Large deviations therefore contribute more than small deviations. Variance is
zero only when all probability is concentrated at one value.

The equivalent computational form is

\[
\operatorname{Var}(X)=E[X^2]-\{E[X]\}^2.
\]

It gives the same result, but the definition
\(E[(X-\mu)^2]\) is usually easier to interpret.

The **standard deviation** is

\[
\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.
\]

Variance has squared units; standard deviation returns to the units of \(X\).
For a Bernoulli variable, \(\operatorname{Var}(X)=p(1-p)\).

```{r b04-variance}
mu_s <- sum(s * p_s)
var_s <- sum((s - mu_s)^2 * p_s)

c(
  Var_X = p_hat * (1 - p_hat),
  SD_X = sqrt(p_hat * (1 - p_hat)),
  Var_S = var_s,
  SD_S = sqrt(var_s)
)
```

#### Distribution variance and sample variance

For observed values \(x_1,\ldots,x_n\), the usual sample variance is

\[
s^2=\frac{1}{n-1}\sum_{i=1}^n(x_i-\bar{x})^2.
\]

If the empirical distribution assigns probability \(1/n\) to each row, its
variance divides by \(n\). R's `var()` uses the conventional sample
denominator \(n-1\).

```{r b04-word-variance}
w <- sharks$description_words
c(
  empirical_variance_divide_n = mean((w - mean(w))^2),
  sample_variance_divide_n_minus_1 = var(w),
  sample_sd = sd(w)
)
```

These are different summaries of the same 706 observations. Calling the sample
variance an estimate of population variance additionally requires a sampling
model.

### Conditional probability

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/conditional-probabilities/">02.2 Conditional Probabilities</a>.</p>

For events \(A\) and \(B\) with \(P(B)>0\),

\[
P(A\mid B)=\frac{P(A\cap B)}{P(B)}.
\]

Conditioning keeps only outcomes inside \(B\), rescales their total probability
to 1, and then asks which of those outcomes are also in \(A\). For a fixed
\(B\), \(P(\cdot\mid B)\) is itself a probability law on the same sample space.

From the table:

\[
P(D\mid L)=\frac{256}{444}\approx0.577,
\qquad
P(L\mid D)=\frac{256}{383}\approx0.668.
\]

The numerator is 256 in both calculations. The denominators—and therefore the
questions—differ.

```{r b04-conditionals}
c(
  P_deal_given_late = mean(D[L]),
  P_late_given_deal = mean(L[D])
)
```

A 66.8% figure refers to \(P(L\mid D)\), not \(P(D\mid L)\): 66.8% of deal
records are from Seasons 5--8, while 57.7% of Seasons 5--8 records reached an
on-air deal.

### The multiplication rule

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-multiplication-rule/">02.6 The Multiplication Rule</a>.</p>

Rearranging the definition gives

\[
P(A\cap B)=P(A\mid B)P(B).
\]

Reversing the roles of \(A\) and \(B\) gives the same joint probability:

\[
P(A\cap B)=P(A\mid B)P(B)=P(B\mid A)P(A).
\]

For the late-season deal event:

```{r b04-multiplication}
c(
  direct_joint = mean(D & L),
  conditional_times_base = mean(D[L]) * mean(L),
  reverse_conditional_times_base = mean(L[D]) * mean(D)
)
```

This is the algebraic reason the two conditional probabilities can share a
numerator while having different denominators.

### Partitions and the law of total probability

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/total-probability-theorem/">02.7 Total Probability Theorem</a>.</p>

Events \(B_1,\ldots,B_k\) form a **partition** of \(\Omega\) when they are
mutually disjoint and their union is \(\Omega\). Every outcome belongs to
exactly one part.

If \(P(B_i)>0\), then:

\[
P(A)=\sum_{i=1}^{k}P(A\mid B_i)P(B_i).
\]

The early/late events \(\{L^c,L\}\) form a partition, so

\[
P(D)=P(D\mid L^c)P(L^c)+P(D\mid L)P(L).
\]

```{r b04-total-probability}
c(
  direct = mean(D),
  from_periods = mean(D[!L]) * mean(!L) + mean(D[L]) * mean(L)
)
```

The overall deal probability is a weighted average of the period-specific deal
probabilities. The group probabilities supply the weights.

### Bayes' rule

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/bayes-rule/">02.8 Bayes' Rule</a>.</p>

From the multiplication rule,

\[
P(B_i\mid A)=\frac{P(A\mid B_i)P(B_i)}{P(A)}.
\]

Using total probability to expand the denominator:

\[
P(B_i\mid A)=
\frac{P(A\mid B_i)P(B_i)}
{\sum_jP(A\mid B_j)P(B_j)}.
\]

For the Shark Tank events:

```{r b04-bayes}
bayes_late_given_deal <- mean(D[L]) * mean(L) / mean(D)
c(
  from_bayes = bayes_late_given_deal,
  from_table = mean(L[D])
)
```

Bayes' rule reverses the direction of conditioning. It does not reverse causal
direction and does not make an observational comparison causal.

#### Base rates and classification

Suppose an AI classifier flags a pitch as likely to receive a deal. To find

\[
P(\text{deal}\mid\text{positive flag}),
\]

we need:

- the deal base rate \(P(\text{deal})\);
- sensitivity \(P(\text{positive}\mid\text{deal})\);
- the false-positive rate
  \(P(\text{positive}\mid\text{no deal})\).

High sensitivity alone is insufficient: the deal base rate and false-positive
rate also determine the proportion of flagged pitches that receive a deal.

:::

::: {.course-phase .phase-after}

## After class

### Retrieval

1. What does variance average?
2. Why do variance and standard deviation have different units?
3. What is the denominator of \(P(A\mid B)\)?
4. Why is \(P(A\mid B)\) generally different from \(P(B\mid A)\)?
5. What makes a collection of events a partition?
6. What does the denominator in Bayes' rule calculate?

<details>
<summary>Check your answers</summary>

1. Variance averages squared deviations from the distribution mean.
2. Variance has squared units; standard deviation is its square root and has
   the variable's original units.
3. \(P(B)\), the probability of the conditioning event.
4. They condition on different reference classes and usually have different
   denominators.
5. The parts are mutually disjoint and their union is \(\Omega\).
6. It calculates the total probability of the observed evidence across all
   mutually exclusive cases in the partition.

</details>

### Practice

1. A variable \(Y\) has PMF \(p_Y(0)=0.2\), \(p_Y(1)=0.5\), and
   \(p_Y(2)=0.3\). Use \(E[Y]=1.1\) and \(E[Y^2]=1.7\) to calculate
   \(\operatorname{Var}(Y)\) and \(\operatorname{SD}(Y)\).
2. Use the early/late partition to verify
   \(P(D)=P(D\mid L^c)P(L^c)+P(D\mid L)P(L)\).
3. A classifier has a 10% deal base rate, 80% sensitivity, and a 20%
   false-positive rate. Calculate \(P(\text{deal}\mid\text{positive flag})\).
4. Calculate and interpret \(P(L\mid D)\). Explain why it differs from
   \(P(D\mid L)\).
5. R reports an empirical variance of 556.365 and a sample variance of
   557.154 for written-description length. Explain why the values differ.

<details>
<summary>Check your answers</summary>

1. \(\operatorname{Var}(Y)=1.7-1.1^2=0.49\) and
   \(\operatorname{SD}(Y)=\sqrt{0.49}=0.7\).
2. \((127/262)(262/706)+(256/444)(444/706)=383/706\approx0.542\).
3. Bayes' rule gives
   \[
   \frac{0.80(0.10)}{0.80(0.10)+0.20(0.90)}
   =\frac{0.08}{0.26}\approx0.308.
   \]
   About 30.8% of positively flagged cases are deals under this model.
4. \(P(L\mid D)=256/383\approx0.668\): among deal records, 66.8% came from
   Seasons 5--8. It differs from \(P(D\mid L)=256/444\approx0.577\) because
   the conditioning event—and therefore the denominator—is different.
5. The empirical-distribution variance weights every one of the 706 records by
   \(1/706\) and therefore divides by \(n\). The usual sample variance divides
   by \(n-1\), so it is slightly larger.

</details>

### Worked exam interpretation

**Question.** A student sees \(P(D\mid L)=0.577\) and writes, “57.7% of deal
pitches came from late seasons.”

<details>
<summary>Check a model answer</summary>

The calculation conditions on \(L\), so the denominator is
the 444 pitches in Seasons 5-8. The correct interpretation is that 57.7% of
late-season records reached an on-air deal. The reverse conditional is
\(P(L\mid D)=256/383=66.8\%\). Neither quantity measures completed investment
or a causal season effect.

</details>

Next: [Lecture 5: Joint random variables, covariance, and correlation](#b05).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 7--8.

:::
