# Lecture 8: Sampling distributions and the central limit theorem {#b08}

```{r b08-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
p_deal <- mean(sharks$deal_on_show)
```

<div class="outcomes">
<span class="label">By the end of Lecture 8, you can</span>
<ul>
<li>distinguish a population distribution from a sampling distribution;</li>
<li>derive the mean and variance of a sample mean;</li>
<li>calculate and interpret a standard error;</li>
<li>explain why estimating population spread leads to a t distribution with \(n-1\) degrees of freedom;</li>
<li>apply the central limit theorem to means and proportions;</li>
<li>explain how sample size changes precision.</li>
</ul>
</div>

::: {.course-phase .phase-before}

## Before class

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch <a href="https://www.youtube.com/watch?v=Yh5bR7X3ch8">The Weak Law
of Large Numbers</a> and <a href="https://www.youtube.com/watch?v=IrKUM3nNXJE">The
Central Limit Theorem</a>. Focus on what converges as sample size grows and on
which distribution becomes approximately normal.
</div>

If square roots, sums, degrees of freedom, or standardisation are not secure,
review [Mathematics Refresher](#math-refresher).

Classify each object:

1. the distribution of equity offered across individual pitches;
2. the distribution of \(\bar X\) across all possible samples of 40 pitches;
3. the single mean calculated from one observed sample.

<details>
<summary>Check your preparation</summary>

The first is a population distribution, the second is a sampling distribution,
and the third is a realised estimate.

</details>

:::

::: {.course-phase .phase-in-class}

## In class

<p class="concept-video"><strong>MIT explanations (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-weak-law-of-large-numbers/">18.4 The Weak Law of Large Numbers</a> and
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-central-limit-theorem/">19.2 The Central Limit Theorem</a>.</p>

### A statistic has a distribution

Imagine repeatedly drawing random samples of the same size and calculating the
sample mean each time. The resulting distribution of possible values of
\(\bar X\) is its **sampling distribution**. It describes uncertainty in an
estimator, not variation among individual pitches.

For independent, identically distributed variables
\(X_1,\ldots,X_n\) with population mean \(\mu\) and variance \(\sigma^2\),

\[
E[\bar X]=\mu,
\qquad
\operatorname{Var}(\bar X)=\frac{\sigma^2}{n},
\qquad
\operatorname{SD}(\bar X)=\frac{\sigma}{\sqrt n}.
\]

The first equality says that \(\bar X\) is **unbiased** for \(\mu\). The third
quantity is the standard deviation of the estimator across repeated samples.
It is called its **standard error** when it is known or estimated from data.

In practice \(\sigma\) is usually unknown, so we estimate it with the sample
SD \(S\):

\[
\operatorname{SE}(\bar X)=\frac{S}{\sqrt n}.
\]

The population SD \(\sigma\) is a fixed parameter. The sample SD \(S\) changes
from sample to sample. Substituting \(S\) therefore adds uncertainty that a
normal reference curve does not fully represent in a small sample.

### From the standard normal to the t distribution

For independent observations from a normal population,

\[
T=\frac{\bar X-\mu}{S/\sqrt n}
\sim t_{n-1}.
\]

The t distribution is symmetric around zero like the standard normal, but it
has heavier tails. Its **degrees of freedom** determine how much additional
tail area is needed:

- few degrees of freedom mean that \(S\) is estimated imprecisely, so the
  tails are heavier;
- more degrees of freedom mean more information about spread;
- as degrees of freedom increase, the t distribution approaches \(N(0,1)\).

For a one-sample mean, \(df=n-1\) because estimating the sample mean uses one
degree of freedom before the sample variance is calculated.

```{r b08-t-distributions, echo=FALSE}
t_grid <- seq(-4, 4, length.out = 800)
plot(
  t_grid,
  dt(t_grid, df = 4),
  type = "l",
  lwd = 3,
  col = "#D55E00",
  bty = "n",
  xlab = "Standardised value",
  ylab = "Density",
  ylim = c(0, 0.41)
)
lines(t_grid, dt(t_grid, df = 10), lwd = 2.5,
      col = "#0072B2", lty = 2)
lines(t_grid, dnorm(t_grid), lwd = 2.5,
      col = "#17324D", lty = 3)
legend(
  "topright",
  c("t, df = 4", "t, df = 10", "standard normal"),
  col = c("#D55E00", "#0072B2", "#17324D"),
  lwd = c(3, 2.5, 2.5),
  lty = c(1, 2, 3),
  bty = "n"
)
```

The difference is most visible in the tails. This is why a 95% t critical
value is larger than 1.96 when \(df\) is small.

### Why averaging reduces variation

For independent observations,

\[
\operatorname{Var}\!\left(\frac{X_1+\cdots+X_n}{n}\right)
=\frac{1}{n^2}\sum_{i=1}^{n}\operatorname{Var}(X_i)
=\frac{\sigma^2}{n}.
\]

Independent positive and negative deviations partly cancel in an average.
Doubling \(n\) does not halve the SE: because of the square root, four times as
many observations are needed to halve it.

### The central limit theorem

The central limit theorem (CLT) says that, under independent sampling with
finite variance, the standardised sample mean

\[
Z=\frac{\bar X-\mu}{\sigma/\sqrt n}
\]

approaches a standard normal distribution as \(n\) grows. Equivalently,

\[
\bar X\ \dot\sim\ N\!\left(\mu,\frac{\sigma^2}{n}\right)
\]

for a sufficiently large sample. The dot above \(\sim\) signals an
approximation.

There is no universal rule that \(n=30\) is always enough. Strong skewness,
extreme outliers, or dependence may require much more data. If the population
itself is normal, the sample mean is normal for every \(n\).

```{r b08-clt-simulation, echo=FALSE}
set.seed(208)
B <- 5000
means_5 <- replicate(B, mean(sample(sharks$equity_offered_pct, 5,
                                    replace = TRUE)))
means_40 <- replicate(B, mean(sample(sharks$equity_offered_pct, 40,
                                     replace = TRUE)))

all_means <- c(means_5, means_40)
shared_limits <- c(floor(min(all_means)), ceiling(max(all_means)))
shared_breaks <- seq(shared_limits[1], shared_limits[2], length.out = 36)
hist_5 <- hist(means_5, breaks = shared_breaks, plot = FALSE)
hist_40 <- hist(means_40, breaks = shared_breaks, plot = FALSE)
shared_ymax <- 1.08 * max(
  hist_5$density,
  hist_40$density,
  dnorm(mu_equity, mu_equity, sigma_equity / sqrt(5)),
  dnorm(mu_equity, mu_equity, sigma_equity / sqrt(40))
)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.4, 4.2, 1.4, 0.8), las = 1)
hist(means_5, breaks = shared_breaks, probability = TRUE, col = "#DCEAF3",
     border = "white", bty = "n", main = "n = 5",
     xlab = "Sample mean equity offered (%)", ylab = "Density",
     xlim = shared_limits, ylim = c(0, shared_ymax))
curve(dnorm(x, mu_equity, sigma_equity / sqrt(5)), add = TRUE,
      col = "#D55E00", lwd = 3)
hist(means_40, breaks = shared_breaks, probability = TRUE, col = "#CDEBDD",
     border = "white", bty = "n", main = "n = 40",
     xlab = "Sample mean equity offered (%)", ylab = "Density",
     xlim = shared_limits, ylim = c(0, shared_ymax))
curve(dnorm(x, mu_equity, sigma_equity / sqrt(40)), add = TRUE,
      col = "#D55E00", lwd = 3)
par(old_par)
```

Both panels use the **same horizontal and vertical scales**. The horizontal
axis shows possible sample means in equity percentage points; the vertical
axis is density. The bars are the simulated sampling distributions and the
orange lines are the corresponding normal approximations. The original
pitch-level variable is strongly right-skewed. Its mean based on 40 independent
draws is nevertheless much more concentrated and closer to a normal shape than
a mean based on 5 draws.

### Sample proportions are sample means

For a Bernoulli variable \(X\) with \(P(X=1)=p\),

\[
E[X]=p,
\qquad
\operatorname{Var}(X)=p(1-p).
\]

Therefore the sample proportion \(\widehat P=\bar X\) satisfies

\[
E[\widehat P]=p,
\qquad
\operatorname{SE}(\widehat P)
=\sqrt{\frac{p(1-p)}{n}}.
\]

Its sampling distribution is approximately normal when the expected numbers
of successes and failures are sufficiently large. This course uses
\(np\ge10\) and \(n(1-p)\ge10\) as the working rule.

### Law of large numbers and CLT answer different questions

- The law of large numbers says \(\bar X\) becomes close to \(\mu\) as \(n\)
  grows.
- The CLT describes the approximate shape and scale of the sampling
  distribution after standardisation.

Neither result repairs selection bias or dependence. More observations reduce
random sampling error; they do not automatically make the data representative.

```{r b08-se-comparison}
c(
  SE_mean_n_25 = sigma_equity / sqrt(25),
  SE_mean_n_100 = sigma_equity / sqrt(100),
  SE_deal_proportion_n_100 = sqrt(p_deal * (1 - p_deal) / 100)
)
```

Increasing the mean sample size from 25 to 100 halves its standard error.

:::

::: {.course-phase .phase-after}

## After class

### Retrieval

1. What varies in a sampling distribution?
2. Why is \(E[\bar X]=\mu\) called unbiasedness?
3. What is the difference between an SD and an SE?
4. Why does estimating \(\sigma\) lead to a t rather than normal reference
   distribution?
5. What does the CLT say becomes approximately normal?
6. Why does a larger sample not remove selection bias?

<details>
<summary>Check your answers</summary>

1. The statistic varies across possible repeated samples.
2. Its long-run average equals the population parameter.
3. An SD describes variation among observations; an SE describes variation of
   an estimator across samples.
4. Replacing \(\sigma\) with the sample SD \(S\) adds sampling uncertainty.
   The t distribution represents that extra uncertainty.
5. The standardised sample mean, and therefore approximately the sampling
   distribution of the mean—not necessarily the original observations.
6. Sample size reduces random error but does not change how units entered the
   sampling frame or sample.

</details>

### Practice

1. If \(\sigma=12\) and \(n=36\), find
   \(E[\bar X]\), \(\operatorname{Var}(\bar X)\), and
   \(\operatorname{SE}(\bar X)\).
2. How large must \(n\) become to reduce an SE from 4 to 1?
3. For \(p=0.30\) and \(n=100\), calculate
   \(\operatorname{SE}(\widehat P)\) and check the success-failure condition.
4. Explain why the normal approximation is more convincing in the right-hand
   simulation panel.
5. State one threat to the CLT-based approximation that a large \(n\) alone
   does not solve.
6. A one-sample mean is based on \(n=25\). State its degrees of freedom and
   explain why the corresponding t critical value exceeds 1.96.

<details>
<summary>Check your answers</summary>

1. \(E[\bar X]=\mu\),
   \(\operatorname{Var}(\bar X)=12^2/36=4\), and
   \(\operatorname{SE}(\bar X)=2\).
2. The SE must fall by a factor of 4, so the sample size must increase by
   \(4^2=16\).
3. \(\sqrt{0.30(0.70)/100}\approx0.0458\). The expected counts are 30 and 70,
   so both exceed 10.
4. Averaging 40 observations reduces skewness and sampling variation more than
   averaging 5 observations.
5. Dependence or a non-representative sampling process.
6. \(df=25-1=24\). The sample SD is estimated after the sample mean has used
   one degree of freedom. The t distribution therefore has heavier tails than
   the standard normal and needs a larger 95% critical value.

</details>

Continue to [Tutorial 3: Sampling and standard errors](#b-ep03).

Formula sheet: [Lecture 8 formulas](#formula-l8).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 13 and 14.

:::
