# Lecture 1: Simple linear regression {#i01}

```{r i01-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_100k <- deals$ask_amount_usd / 100000
deals$deal_100k <- deals$deal_amount_usd / 100000
deal_model <- lm(deal_100k ~ ask_100k, data = deals)
display_deals <- subset(deals, ask_100k <= 10 & deal_100k <= 10)
```

<div class="outcomes">
<span class="label">By the end of Lecture 1, you can</span>
<ul>
<li>state a simple linear regression model and identify its components;</li>
<li>explain least-squares estimation;</li>
<li>distinguish observations, fitted values, errors, and residuals;</li>
<li>fit a regression with <code>lm()</code> in R;</li>
<li>interpret an intercept and slope in units and within the observed range.</li>
</ul>
</div>

::: {.course-phase .phase-before}

## Before class

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=vCHtY6Me5FI&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=9">3.1
Simple Linear Regression</a>. Focus on the population line, the fitted line,
and least squares.
</div>

If lines, slopes, intercepts, or subscripts are not secure, review
[Mathematics Refresher: straight lines and indices](#math-refresher).

For the 882 pitches with an on-air deal, define:

- \(Y_i\): agreed on-air deal amount, in $100,000s;
- \(X_i\): amount initially requested, in $100,000s.

Before fitting a model, predict the sign and plausible units of the slope.

<details>
<summary>Check your preparation</summary>

The slope should be positive: larger requests tend to accompany larger agreed
amounts. Its units are $100,000 of agreed amount per $100,000 requested, which
simplifies to a ratio.

</details>

:::

::: {.course-phase .phase-in-class}

## In class

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=vCHtY6Me5FI">3.1 Simple Linear Regression</a>.</p>

### Begin with the data

```{r i01-scatter, echo=FALSE}
plot(
  display_deals$ask_100k,
  display_deals$deal_100k,
  pch = 16,
  col = grDevices::adjustcolor("#0072B2", alpha.f = 0.30),
  bty = "n",
  xlab = "Amount requested ($100,000s)",
  ylab = "Agreed on-air deal amount ($100,000s)",
  xlim = c(0, 10.3),
  ylim = c(0, 10.3),
  xaxs = "i",
  yaxs = "i"
)
abline(deal_model, lwd = 2.2,
       col = grDevices::adjustcolor("#D55E00", alpha.f = 0.78))
legend("topleft", "Fitted regression", col = "#D55E00", lwd = 2.2,
       bty = "n")
```

The display shows 864 of the 882 deal records so that the dense region is
readable. Eighteen records above $1 million on at least one axis are omitted
from this display but retained when estimating the orange line. Always say
when a graph is zoomed; otherwise readers may mistake the displayed range for
the complete data.

### The population model

The simple linear regression model is

\[
Y_i=\beta_0+\beta_1X_i+\varepsilon_i.
\]

Equivalently, the conditional mean is

\[
E[Y_i\mid X_i=x]=\beta_0+\beta_1x.
\]

- \(Y_i\) is the outcome for observation \(i\).
- \(X_i\) is the predictor.
- \(\beta_0\) is the population intercept.
- \(\beta_1\) is the population slope.
- \(\varepsilon_i\) is the part of \(Y_i\) not represented by the line.

The slope is

\[
\beta_1
=E[Y\mid X=x+1]-E[Y\mid X=x].
\]

It is the expected change in \(Y\) associated with a one-unit increase in
\(X\). “Associated with” is deliberate: a regression line alone does not
identify a causal effect.

### Least-squares estimation

The fitted sample line is

\[
\widehat Y_i=b_0+b_1X_i.
\]

For each observation, the residual is the vertical difference

\[
e_i=Y_i-\widehat Y_i.
\]

#### Why define error vertically?

For observation \(i\), the model predicts \(\widehat Y_i\) at its observed
\(X_i\). The residual \(e_i=Y_i-\widehat Y_i\) is therefore a difference in
the **outcome's units**. A positive residual lies above the line; a negative
residual lies below it.

Adding raw residuals is not a useful score: positive and negative misses can
cancel. Adding absolute residuals avoids cancellation, but the resulting
objective has sharp corners and is harder to optimise algebraically. Squared
residuals are non-negative, penalise large misses more strongly, and give a
smooth objective with a tractable solution. This is a modelling choice, not a
claim that squared error is the only possible loss function.

Ordinary least squares chooses \(b_0\) and \(b_1\) to minimise

\[
\operatorname{SSE}=\sum_{i=1}^{n}e_i^2
=\sum_{i=1}^{n}(Y_i-b_0-b_1X_i)^2.
\]

The initials **OLS** stand for **ordinary least squares**: the rule that fits
the linear model by minimising this squared-error criterion.

#### Try the criterion yourself

Drag either square handle. The vertical segments are residuals; the displayed
SSE updates after every move. Then reveal the line with the smallest SSE.

<div id="least-squares-lab" class="ls-lab" role="group" aria-labelledby="ls-lab-title">
<div class="ls-lab-heading">
<strong id="ls-lab-title">Least-squares line lab</strong>
<span id="ls-equation" aria-live="polite"></span>
</div>
<svg id="ls-svg" viewBox="0 0 700 390" role="img" aria-label="Scatterplot with a draggable line and vertical residual segments"></svg>
<div class="ls-controls">
<label>Line height at x = 1
<input id="ls-left" type="range" min="0" max="10" step="0.05" value="3">
</label>
<label>Line height at x = 9
<input id="ls-right" type="range" min="0" max="10" step="0.05" value="7">
</label>
</div>
<div class="ls-actions">
<button id="ls-best" type="button">Show least-squares fit</button>
<button id="ls-reset" type="button">Try another line</button>
<strong id="ls-sse" aria-live="polite"></strong>
</div>
<p class="ls-caption">Squares make every contribution to SSE non-negative. Longer residuals contribute disproportionately because their lengths are squared.</p>
</div>

<script>
(function () {
  const root = document.getElementById("least-squares-lab");
  if (!root || root.dataset.ready) return;
  root.dataset.ready = "true";
  const svg = root.querySelector("#ls-svg");
  const left = root.querySelector("#ls-left");
  const right = root.querySelector("#ls-right");
  const equation = root.querySelector("#ls-equation");
  const sseText = root.querySelector("#ls-sse");
  const points = [
    {x:1, y:1.8}, {x:2, y:1.4}, {x:3, y:3.2},
    {x:4, y:4.8}, {x:5, y:4.2}, {x:6, y:6.8},
    {x:7, y:6.4}, {x:8, y:8.6}, {x:9, y:8.2}
  ];
  const ns = "http://www.w3.org/2000/svg";
  const box = {left:58, right:680, top:18, bottom:336};
  const sx = x => box.left + x / 10 * (box.right - box.left);
  const sy = y => box.bottom - y / 10 * (box.bottom - box.top);
  const add = (name, attrs, parent=svg) => {
    const el = document.createElementNS(ns, name);
    Object.entries(attrs).forEach(([k,v]) => el.setAttribute(k, v));
    parent.appendChild(el); return el;
  };
  add("line", {x1:box.left, y1:box.bottom, x2:box.right, y2:box.bottom, class:"ls-axis"});
  add("line", {x1:box.left, y1:box.top, x2:box.left, y2:box.bottom, class:"ls-axis"});
  [0,2,4,6,8,10].forEach(v => {
    add("line", {x1:sx(v), y1:box.bottom, x2:sx(v), y2:box.bottom+6, class:"ls-axis"});
    const tx=add("text", {x:sx(v), y:box.bottom+25, class:"ls-tick"}); tx.textContent=v;
    add("line", {x1:box.left-6, y1:sy(v), x2:box.left, y2:sy(v), class:"ls-axis"});
    const ty=add("text", {x:box.left-12, y:sy(v)+4, class:"ls-tick", "text-anchor":"end"}); ty.textContent=v;
  });
  const xl=add("text", {x:(box.left+box.right)/2, y:382, class:"ls-label", "text-anchor":"middle"}); xl.textContent="Predictor x";
  const yl=add("text", {x:16, y:(box.top+box.bottom)/2, class:"ls-label", transform:`rotate(-90 16 ${(box.top+box.bottom)/2})`, "text-anchor":"middle"}); yl.textContent="Outcome y";
  const residualGroup=add("g", {class:"ls-residuals"});
  const line=add("line", {class:"ls-fit-line"});
  const handleLeft=add("rect", {width:16, height:16, rx:2, class:"ls-handle", "aria-hidden":"true"});
  const handleRight=add("rect", {width:16, height:16, rx:2, class:"ls-handle", "aria-hidden":"true"});
  points.forEach(p => add("circle", {cx:sx(p.x), cy:sy(p.y), r:5.5, class:"ls-point"}));
  const meanX=points.reduce((a,p)=>a+p.x,0)/points.length;
  const meanY=points.reduce((a,p)=>a+p.y,0)/points.length;
  const bestSlope=points.reduce((a,p)=>a+(p.x-meanX)*(p.y-meanY),0) /
                  points.reduce((a,p)=>a+(p.x-meanX)**2,0);
  const bestIntercept=meanY-bestSlope*meanX;
  const fitFromHeights = () => {
    const y1=Number(left.value), y9=Number(right.value);
    const slope=(y9-y1)/8, intercept=y1-slope;
    return {intercept, slope};
  };
  function draw() {
    const {intercept, slope}=fitFromHeights();
    residualGroup.replaceChildren();
    let sse=0;
    points.forEach(p => {
      const fitted=intercept+slope*p.x;
      sse+=(p.y-fitted)**2;
      add("line", {x1:sx(p.x), y1:sy(p.y), x2:sx(p.x), y2:sy(fitted), class:"ls-residual"}, residualGroup);
    });
    line.setAttribute("x1", sx(0)); line.setAttribute("y1", sy(intercept));
    line.setAttribute("x2", sx(10)); line.setAttribute("y2", sy(intercept+10*slope));
    [[handleLeft,1,Number(left.value)],[handleRight,9,Number(right.value)]].forEach(([h,x,y])=>{
      h.setAttribute("x",sx(x)-8); h.setAttribute("y",sy(y)-8);
    });
    equation.textContent=`ŷ = ${intercept.toFixed(2)} ${slope<0?"−":"+"} ${Math.abs(slope).toFixed(2)}x`;
    sseText.textContent=`SSE = ${sse.toFixed(2)}`;
  }
  function drag(handle, slider) {
    handle.addEventListener("pointerdown", event => {
      handle.setPointerCapture(event.pointerId);
      const move = e => {
        const rect=svg.getBoundingClientRect();
        const svgY=(e.clientY-rect.top)*390/rect.height;
        const y=Math.max(0,Math.min(10,(box.bottom-svgY)*10/(box.bottom-box.top)));
        slider.value=y.toFixed(2); draw();
      };
      handle.addEventListener("pointermove", move);
      handle.addEventListener("pointerup", () => handle.removeEventListener("pointermove", move), {once:true});
    });
  }
  drag(handleLeft,left); drag(handleRight,right);
  [left,right].forEach(control => control.addEventListener("input",draw));
  root.querySelector("#ls-best").addEventListener("click",()=>{
    left.value=(bestIntercept+bestSlope).toFixed(2);
    right.value=(bestIntercept+9*bestSlope).toFixed(2); draw();
  });
  root.querySelector("#ls-reset").addEventListener("click",()=>{
    left.value=3; right.value=7; draw();
  });
  draw();
})();
</script>

For simple regression,

\[
b_1=\frac{s_{XY}}{s_X^2},
\qquad
b_0=\bar y-b_1\bar x.
\]

The fitted line therefore passes through \((\bar x,\bar y)\).

### Fit and read the model in R

```{r i01-fit}
deal_model <- lm(deal_100k ~ ask_100k, data = deals)
coef(deal_model)
```

The fitted equation is approximately

\[
\widehat{\text{deal amount}}
=0.276+1.009\times\text{amount requested},
\]

with both monetary variables measured in $100,000s.

The slope means that one additional $100,000 requested is associated with
about $100,900 more in the fitted agreed amount among pitches that reached an
on-air deal. It does not mean that asking for more causes investors to offer
more.

The intercept predicts about $27,600 at a request of zero. A zero request lies
outside the observed range and is not substantively meaningful; an intercept
can be mathematically necessary without having a useful standalone
interpretation.

```{r i01-fitted-residual}
head(data.frame(
  observed = deals$deal_100k,
  fitted = fitted(deal_model),
  residual = residuals(deal_model)
))
```

For every row, `observed = fitted + residual`.

### Prediction stays inside the evidence boundary

```{r i01-one-prediction}
predict(deal_model, newdata = data.frame(ask_100k = 3))
```

For a $300,000 request, the fitted mean agreed amount is about $330,000. This
is a model-based average, not a guarantee for a new pitch. Extrapolating far
beyond the observed request range would rely almost entirely on the assumed
line.

:::

::: {.course-phase .phase-after}

## After class

### Retrieval

1. What is the difference between \(\varepsilon_i\) and \(e_i\)?
2. What does least squares minimise?
3. Why must a slope interpretation include units?
4. Why can an intercept be mathematically valid but substantively unhelpful?
5. Why is the slope not automatically causal?

<details>
<summary>Check your answers</summary>

1. \(\varepsilon_i\) is an unobserved population-model error; \(e_i\) is the
   observed residual after fitting the sample line.
2. The sum of squared residuals.
3. The slope is change in outcome units per one predictor unit.
4. The value \(X=0\) may lie outside the observed range or lack substantive
   meaning.
5. The predictor was not randomly assigned and other factors may affect both
   variables.

</details>

### Practice

1. Use the fitted equation to predict the agreed amount for requests of
   $100,000 and $500,000.
2. Calculate the residual for a pitch that requested $300,000 and agreed
   $250,000.
3. Explain why `lm(y ~ x, data = df)` treats the variable left of `~` as the
   outcome.
4. Fit `equity_offered_pct ~ season` and interpret both coefficients.
5. State the unit of every quantity in Question 4.

<details>
<summary>Check your answers</summary>

1. Substitute `ask_100k = 1` and `ask_100k = 5`; the fitted values are about
   1.284 and 5.318 in $100,000s.
2. The fitted amount is about 3.301 in $100,000s, so the residual is
   \(2.5-3.301\approx-0.801\), or about \(-\$80{,}100\).
3. R formulas use `outcome ~ predictors`.
4. The fitted intercept is about 20.46 percentage points and the slope about
   \(-0.824\) percentage points per season. Later seasons are associated with
   smaller initial equity offers on average.
5. The outcome and intercept use percentage points; the predictor uses seasons;
   the slope uses percentage points per season.

</details>

Continue to [Lecture 2: Regression assumptions and coefficient inference](#i02).

For a more formal treatment, see James et al., *An Introduction to Statistical
Learning*, 2nd ed., Sections 3.1 and 3.6.2.

:::
